Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage

### 2023-10-16 ms/ms.Rmd
edits to ms and model table

```{r}
source("r/header.R")
```

@@

*Arabidopsis thaliana* plants were grown in three different light environments: low light (50 PAR), medium light (100 PAR), and high light (200 PAR)... Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using xxx bifocal microscope. This microscope allows the capture of two focal lengths that are spatially correlated such that pixel (1,1) of the abaxial surface image is directly above pixel (1,1) of the adaxial surface image, and so on... Images were 512 by 512 pixels and covered X leaf surface area. We captured 132 images in total, making 66 abaxial-adaxial image pairs. The position of all stomata were recorded using ImageJ [@schneider_nih_2012].

@@

We tested whether stomata are non-randomly distributed by comparing the observed stomatal patterning to a random uniform pattern. For each leaf surface image with $n$ stomata we generated $10^3$ synthetic surfaces with $n$ stomata uniformly randomly distributed on the surface. For each sample image, we compared the observed Nearest Neighbor Index ($\mathrm{NNI}$) to the null distribution of $\mathrm{NNI}$ values calculated from the synthetic data set. The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).

$\mathrm{NNI}$ is the ratio of observed mean distance ($\overline{D}_O$) to the expected mean distance ($\overline{D}_E$) where $\overline{D}_E$ is:

\begin{equation}\label{eq:emd}
  \overline{D}_E = \frac{0.5}{\sqrt{A_\text{leaf} / n_\text{stomata}}}
\end{equation}

\noindent $A_\text{leaf}$ is leaf area visible in the sampled field and $n_\text{stomata}$ the number of stomata. $\overline{D}_E$ is the theoretical average distance to the nearest neighbor of each stomate if stomata were uniformly randomly distributed [@clark_distance_1954]. And $\overline{D}_O$ calculated for each synthetic data set is:

\begin{equation}\label{eq:omd}
  \overline{D}_O = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}}
\end{equation}

$d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023].

@@

\noindent $\mathrm{NNI}$ is calculated for each sample image as described above; $\text{median}(\mathrm{NNI_{random}})$ and $\text{median}(\mathrm{NNI_{uniform}})$ are calculated from the dry og synthetic data specific to each sample image as described above. We tested whether light treatment affects $\mathrm{DI}$ and $D_S$ using analysis of variance (ANOVA).

@@

Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model with the R package **brms** version `r packageVersion("brms")` [@burkner_brms_2017; @burkner_advanced_2018]. Stomatal zone area was calculated using Voronoi tessellation (e.g. Fig. \ref{fig:example}). The stomatal zone area, $S_\text{area}$, is the region of the leaf surface whose distance to stomate, $S$, is less than the distance to any other stomate, $S$. Stomatal length was measured in `ImageJ` [@schneider_nih_2012].

@@

To test whether the position of ab- and adaxial stomata are coordinated we compared the observed distribution to a null distribution where the positions on each surface are random. For each pair of surfaces (observed or synthetic) we calculated the distance squared between each to the nearest stomatal centroid with the R package **raster** version `r packageVersion("raster")`. Then we calculated the cell-wise Pearson correlation coefficient. If stomatal positions on each surface are coordinated to minimize the distance between mesophyll and the nearest stomate, then we expect a negative correlation. A cell that is far from a stomate on one surface should be near a stomate on the other surface (Fig. \ref{fig:amphi}). We generated a null distribution of the correlation coefficient by simulating $10^3$ synthetic data sets for each observed pair. For each synthetic data set, we simulated stomatal position using a random uniform distribution, as described above, matching the number of stomata on abaxial and adaxial leaf surfaces. Stomatal positions on each surface are coordinated if the correlation coefficient is greater than 95% of the synthetic correlation values (one-tailed test).

@@

We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation (Parkhurst 1994). Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$. The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface. With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness. We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate). The advantage of offsetting stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:

@@

We modelled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage.

@@

[NOTE: we should probably adjust light attenuation to be proportional to the chlorophyll concentration which is one of the parameters in the biochemical models.]

@@

FYI - something is wrong with the model. I think the units are off somewhere because the result is very sensitive to grid size and the values often aren't reasonable. I'll need to troubleshoot more, but maybe something will be obvious to you.

We used the `steady.2D()` function in the R package **rootSolve**. The function needs a function to calculate the time differential for node $ij$, $\frac{dC_{m,ij}}{dt}$. The R function below uses the light propagation, biochemical, and diffusional model to calculate a matrix of $\frac{dC_{m,ij}}{dt}$ values:

>>>>

```{r}
source("r/header.R")
fit = readr::read_rds("objects/fit.rds")
```

@@

*Arabidopsis thaliana* plants were grown in three different light environments: low light (PAR = 50 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), medium light (100 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), and high light (200 $\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$). PAR stands for photosynthetically active radiation. Once leaves were mature, we captured images of the abaxial and adaxial leaf surfaces using XXX. We captured 132 images in total, making 66 abaxial-adaxial image pairs. We measured stomatal position and size using ImageJ [@schneider_nih_2012].

@@

We tested whether stomata are non-randomly distributed by comparing the observed stomatal patterning to a random uniform pattern. For each leaf surface image with $n$ stomata we generated $10^3$ synthetic surfaces with $n$ stomata uniformly randomly distributed on the surface. For each sample image, we compared the observed Nearest Neighbor Index ($\mathrm{NNI}$) to the null distribution of $\mathrm{NNI}$ values calculated from the synthetic data set. $\mathrm{NNI}$ is the ratio of observed mean distance ($\overline{D}_O$) to the expected mean distance ($\overline{D}_E$) where $\overline{D}_E$ is:

\begin{equation}\label{eq:emd}
  \overline{D}_E = \frac{0.5}{\sqrt{A_\text{leaf} / n_\text{stomata}}}.
\end{equation}

\noindent $A_\text{leaf}$ is leaf area visible in the sampled field and $n_\text{stomata}$ the number of stomata. $\overline{D}_E$ is the theoretical average distance to the nearest neighbor of each stomate if stomata were uniformly randomly distributed [@clark_distance_1954]. $\overline{D}_O$ calculated for each synthetic data set is:

\begin{equation}\label{eq:omd}
  \overline{D}_O = \frac{\sum_{i=1}^{n_\text{stomata}}d_i}{n_\text{stomata}},
\end{equation}

\noindent where $d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023]. The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test).

@@

\noindent $\mathrm{NNI}$ is calculated for each sample image as described above; $\text{median}(\mathrm{NNI_{random}})$ and $\text{median}(\mathrm{NNI_{uniform}})$ are calculated from the synthetic data specific to each sample image as described above. We tested whether light treatment affects $\mathrm{DI}$ and stomatla density ($D_S$) using analysis of variance (ANOVA).

@@

Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian generalized non-linear multilevel model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023]. Stomatal zone area was calculated using Voronoi tessellation (e.g. Fig. \ref{fig:example}). The stomatal zone area, $S_\text{area}$, is the region of the leaf surface whose distance to stomate, $S$, is less than the distance to any other stomate, $S$. Stomatal length was measured in ImageJ [@schneider_nih_2012].

@@

To test whether the position of ab- and adaxial stomata are coordinated we compared the observed distribution to a null distribution where the positions on each surface are random. For each pair of surfaces (observed or synthetic) we calculated the distance squared between each to the nearest stomatal centroid with the *R* package **raster** version `r packageVersion("raster")`. Then we calculated the cell-wise Pearson correlation coefficient. If stomatal positions on each surface are coordinated to minimize the distance between mesophyll and the nearest stomate, then we expect a negative correlation. A cell that is far from a stomate on one surface should be near a stomate on the other surface (Fig. \ref{fig:amphi}). We generated a null distribution of the correlation coefficient by simulating $10^3$ synthetic data sets for each observed pair. For each synthetic data set, we simulated stomatal position using a random uniform distribution, as described above, matching the number of stomata on abaxial and adaxial leaf surfaces. Stomatal positions on each surface are coordinated if the correlation coefficient is greater than 95% of the synthetic correlation values (one-tailed test).

@@

We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_intercellular_1994] using the finite element method (FEM) following @earles_excess_2017. Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$. The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface. With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness. We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate). The 'coordination advantage' of offsett stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:

@@

We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (Table X).

WORKING HERE - make table of parameter combinations!

@@

@@

We used the `steady.2D()` function in the *R* package **rootSolve** version `r packageVersion("rootSolve")` [@soetaert_practical_2009].
